---
title: "CFAcoop"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_heigth: 9
vignette: >
  %\VignetteIndexEntry{CFAcoop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# How to use this package (and why) 

## Introduction  

The **CFAcoop** package equips you with functions to analyze data from the 
clonogenic assay (also called 'Colony Formation Assay') 
in presence or absence of cellular cooperation. 
Thus, this package allows you to robustly extract clonogenic survival 
information for your cell lines under a given treatment. 

This vignette is meant to enable you to process your data following the method 
first presented in Brix et al., Radiation Oncology, 2020: 
"The clonogenic assay: robustness of plating efficiency-based analysis is 
strongly compromised by cellular cooperation".
Therefore, the data presented in Figure 2 in Brix et al., 
which is provided within the package, 
is used to illustrate how to use the **CFAcoop** package.
Further, it is shown, how the survival 
fractions of cooperative cell lines would look like, if cellular 
cooperation was ignored.

## Cellular Cooperation ##

In order to avoid confusion, cellular cooperation should be defined.
Understanding this concept comes easy, when starting from the opposite.
Clonogenic growth of non-cooperative cells is independent of cell density.
Thus, there is a constant relationship between cells seeded $S$ and colonies 
counted $C$:
\[C = a \cdot S,\]
where $a$ corresponds just to the conventional plating efficiency 
($PE = \frac{C}{S}$).

Now, cellular cooperation refers to the benefit cells can have from their 
surrounding neighbors which results in a non-constant relation of cells seeded 
and colonies counted (For details, see Brix et al., Radiation Oncology, 2020).
The probability of clonogenic growth for a single cell increases with the number
of surrounding cells to cooperate with.
It has turned out that generalizing the equation above by a parameter $b$ 
adequately models the colonies counted of cooperative and non-cooperative cell 
lines:
\[C = a \cdot S^b.\]
In this model, $b = 1$ gives the non-cooperative case and $b > 1$ corresponds to 
cooperative growth.

In short, a cell line is called cooperative, if $b > 1$.

## Clonogenic Survival ##

Conventionally, clonogenic survival at a given treatment $x$ was determined as 
the ratio of colonies counted $C_x$ and the cells seeded $S_x$ 
scaled to the plating efficiency of a reference 
$PE_0 = \frac{C_0}{S_0}$
\[SF'_x = \frac{\frac{C_x}{S_x}}{PE_0}.\]

The new method now shifts the focus of the survival fraction 
directly to the number of cells needed to be seeded under the two conditions 
(treated and untreated) in order to achieve an __identical__ expectation of the 
number of colonies formed $C$.
Essentially, the new method does not focus on the number of colonies formed 
after growth in different cell densities, but on the number of seeded single 
cells with clonogenic potential before growing to identical colony numbers.

\[SF_x(C) = \frac{S_0(C)}{S_x(C)} 
        = exp\left( \frac{log\left(\frac{C}{a_0}\right)}{b_0}
          - \frac{log\left(\frac{C}{a_x}\right)}{b_x}\right)\]

Obviously, for $b_x = b_0 = 1$ the equivalence $SF_x(C) \equiv SF'$ holds for 
all $C$, and thus, the 
non-cooperative case is well covered by the new method. 
Importantly, the conventional determination of clonogenic survival is 
heavily compromised by cellular cooperation, if present.

## Getting Started ##

The data as presented in Figure 2 in Brix et al. 
is included in the package in form of a `` `data.frame` `` *CFAdata*. 
It can be loaded and summarized by: 

```{r loadData}
library(CFAcoop)
data("CFAdata")
summary(CFAdata)
```

# Fast Analysis and Plotting of Results

The shortcut to analyze data, is using the wrapper 
function `` `analyze_survival(RD, name, xtreat)` ``
where RD is a `` `data.frame` `` or `` `matrix` `` containing your numbers of
seeded cells (first column) and numbers of colonies counted under the 
treatments (numeric argument, e.g. the dose 
applied `` `xtreat = c(0,1,2,4,6,8)` ``).

The returned objects should be concatenated in a list-object and can be plotted 
by `` `plot_sf()` ``.

```{r show1, fig.width=7, fig.height=5}
data("CFAdata")
data1 <- subset.data.frame(x = CFAdata, subset = CFAdata$cell.line == "T47D")
data2 <- subset.data.frame(x = CFAdata, subset = CFAdata$cell.line == "BT20")
SF <- vector("list", 2)
SF[[1]] <- analyze_survival(
  RD = data1[, c("Cells seeded","0 Gy","1 Gy","2 Gy","4 Gy","6 Gy","8 Gy")], 
  name = as.character(data1[1,1]), 
  xtreat = c(0, 1, 2, 4, 6, 8), 
  c_range = c(5, 20, 100))
SF[[2]] <- analyze_survival(
  RD = data2[,-1], 
  name = as.character(data2[1,1]), 
  xtreat = c(0, 1, 2, 4, 6, 8))
plot_sf(SF = SF)
```

Raw data from single replicates is plotted as '$+$'-symbols.
Corresponding regression lines are calculated using the mean values of 
replicates of identical numbers of cells seeded, which are plotted as dots.
The color indicates the treatment (irradiation with 0 to 8 Gy) 
and links the numbers of colonies counted 
from the upper plot to the calculated clonogenic survival in the lower plot. 
Shaded areas indicate the span from $C = 5$ to $C = 100$, which is 
within the target region of good experimental practice.
Dashed lines show regression lines with a slope of $b=1$ (at $log10(a)$ 
varying from 0 to 4) for orientation, so that 
any substantially non-linear relation (i.e. $b \neq 1$) between the number of 
colonies counted ($C$) 
and the number of cells seeded ($S$) can be spotted easily.  
The dots in the treatment response curves correspond to the survival fractions
at $C = 5, 20$ and $100$, with error bars indicating the uncertainty bands 
calculated as $2.5\%$ and $97.5\%$-percentiles of the parametric bootstrap 
($N = 1000$) evaluating the covariance matrices of the power regressions.

All information used for plotting is contained in the objects returned by 
`` `analyze_survival` ``. 

A `` `data.frame` `` with a summary of the estimated survival fractions can be 
generated by 
```{r export_sf}
summary_df <- export_sf(SF)
```

to export this `` `data.frame` `` in a csv-File, execute:
`` `
write.csv(x = summary_df,file = "CFAcoopResult.csv")
` ``

The `` `data.frame` `` includes the following columns
```{r summary_export_sf}
colnames(summary_df)
head(summary_df)
summary(summary_df)
```

All information of this `` `data.frame` `` is also accessible directly in the 
object returned by `` `analyze_survival` ``.
For instance, the information about the regression of the 0 Gy reference of the 
cell line BT20 or the survival fractions of the 5 treatments for T47D 
(at $C = 5, 20$ and $100$) can be recalled by: 
```{r sf_details}
SF[[1]]$fit[1]
SF[[2]]$SF
```

# Details for Focussed Analysis

## Assess Cellular Cooperation

Key to the robust analysis of clonogenic analysis data is the modeling of the 
cellular cooperation. 
We assume that the underlying functional dependency of seeded cells and counted 
colonies is of the form
\[C = a \cdot S^{b},\]
where $b$ indicates the degree of cellular cooperation 
($b = 1$ is implicitely assumed for the PE-based approach).

The coefficient $b$ is estimated in a linear regression model
\[log(C) = log(a) + b \cdot log(S) + \varepsilon, \varepsilon \sim \mathcal{N}(0,\sigma^2). \]

The function `` `pwr_reg(seede, counted)` `` provides this regression and 
returns a `` `summary.lm` `` object.

Note that the analysis of cellular cooperation is restricted to the range 
of seeded cells, where at least one colony was observed. 
Outside this range, the attempt of studying clonogenic survival based on no 
observed colony counts is not reasonable and thus, `` `pwr_reg` `` will remove those 
data points from analysis. 
Thus, it is strongly recommended to use the averaged data for regression.
By doing so, the range of the independent variable of the regression is widened.
(Removing only those replicates with no colonies 
at one or few specific cell densities would bias the model fitting.)

```{r PowerReg, fig.width=5, fig.height=4}
data <- subset.data.frame(x = CFAdata, subset = CFAdata$cell.line == "BT20")
data <- aggregate(x = data[, -1], by = list(data$`Cells seeded`), FUN = "mean", na.rm = TRUE)
par_0 <- pwr_reg(seeded = data$`Cells seeded`, counted = data$`0 Gy`)
par_0$coefficients
plot(x = log10(data$`Cells seeded`), y = log10(data$`0 Gy`),xlim = c(2,3.5))
abline(a = log10(exp(1)) * par_0$coefficients[1, 1], b = par_0$coefficients[2, 1])
```

With the results of this function, we can also test for cellular cooperation.
Note, that the _p-value_ in the _coefficients_ table corresponds to the 
null hypothesis $b = 0$, but we are interested in the null hypothesis of 
$b \leq 1$. 
Thus, we find our p-value of interest by computing

```{r TestingCooperation}
p_value <-
  (1 - pt(
    q = (par_0$coefficients[2, 1] - 1) / par_0$coefficients[2, 2],
    df = par_0$df[2]
  ))
```

Thus, BT20 is higly cooperative 
($b = 1.76$, $\hat{\sigma}_b = 0.12$, $p < 0.001$).

## Determine Clonogenic Survival Fractions

In this package, the survival fraction $SF(C)$ for clonogenic survival is 
calculated as the number of cells that need to be seeded without treatment 
divided by the number of cells needed to be seeded with treatment for obtaining 
__the same__ expectation of colonies counted __$C$__.
\[ SF(C) = \frac{S_0(C)}{S_x(C)} = exp\left( \frac{log\left(\frac{C}{a_0}\right)}{b_0} - \frac{log\left(\frac{C}{a_x}\right)}{b_x} \right)\]

Given two parameter sets of clonogenic assay data, the clonogenic survival 
can be calculated as:
```{r calculateSF}
data <- subset.data.frame(x = CFAdata, subset = CFAdata$cell.line == "BT20")
data <- aggregate(x = data[, -1], by = list(data$`Cells seeded`), FUN = "mean", na.rm = TRUE)
par_0 <- pwr_reg(seeded = data$`Cells seeded`, counted = data$`0 Gy`)
par_4 <- pwr_reg(seeded = data$`Cells seeded`, counted = data$`4 Gy`)
calculate_sf(par_ref = par_0, par_treat = par_4, c_range = c(5, 100))
```

Since the SF-values are solely dependent on the estimated parameters in the 
power regression, the inherent uncertainty can be assessed via parametric 
bootstrapping, namely:
```{r bootstrapUncertainty}
ParSet_0 <- mvtnorm::rmvnorm(n = 10^4,mean = par_0$coefficients[,1],
                             sigma = vcov(object = par_0))
ParSet_4 <- mvtnorm::rmvnorm(n = 10^4,mean = par_4$coefficients[,1],
                             sigma = vcov(object = par_4))
parbootstrap <- calculate_sf(par_ref = ParSet_0, 
                             par_treat = ParSet_4, 
                             c_range = c(5, 100))
cat(c(mean(parbootstrap),sd(parbootstrap)))
cat(quantile(x = parbootstrap,probs = c(0.025,0.5,0.975)))
```


### Remark on Ignoring Cellular Cooperation 
Note that in case of cooperative cell lines, the parameter $a$ does not 
correspond to a plating efficiency as for non-cooperative cell lines.
The concept of a characteristic plating efficiency does not apply to 
cooperative cell lines.

# What's the problem with PE-based analysis? 

In short: The plating efficiency frequently is not as constant as it needs to 
be in order to serve as an adequate normalization factor.

To illustrate this, we compare the PE-based calculated $SF'$-values 
with the $SF(C = 20)$-values calculated with the new method for 
(1) the non-cooperative cell line T47D and 
(2) the cooperative cell line (BT20).

## (1) The non-cooperative case

For calculating the survival fraction, plating efficiencies are required.
Plating efficiencies ($C/S$) are calculated easily as:

```{r PEfail1a, fig.width=5, fig.height=4}
data(CFAdata)
data <- subset.data.frame(x = CFAdata, subset = CFAdata$cell.line == "T47D")
data <- aggregate(x = data[, -1], by = list(data$`Cells seeded`), FUN = "mean", na.rm = TRUE)
PE_x <- data$`4 Gy` / data$`Cells seeded`
PE_0 <- data$`0 Gy` / data$`Cells seeded`
plot(x = rep(c(0, 1), each = 18), y = c(PE_0, PE_x), lty = 0, ylim = c(0,0.5),xlim = c(-0.1,1.1),
     xlab = "treatment", ylab = "C / S", axes = FALSE, main = "T47D")
axis(side = 1,at = c(0,1),labels = c("0 Gy","4 Gy"))
axis(side = 2,at = seq(0,0.5,0.1))
```

Now, which values are to be compared? 

When there is no effect of the cell density, as assumed by the conventional 
approach (not taking cellular cooperation into account), 
each combination (of a $C_{0 Gy}/S_{0 Gy}$- and $C_{4 Gy}/S_{4 Gy}$-value) is 
equally reliable. 

Thus, the full set of all combinations is: 

```{r PEfail1b, fig.width=5, fig.height=4}
SF_resample <- rep(PE_x, each = length(PE_0)) / rep(PE_0, times = length(PE_x))
hist(SF_resample, breaks = 25,xlim = c(0.12,0.25),xlab = "(C_4/S_4) / (C_0/S_0)",
     main = "valid PE-based SF'-values")
```

Without the assessment of cellular cooperation, conventional calculation 
of an $SF'$-value, corresponds to picking randomly a sample from the 
distribution shown in the histogram above.

A comparison of the range of this distribution and the calculated uncertainties
of the new method shows, that for non-cooperative cell lines such as T47D, there 
is no big difference in this variability/uncertainty.

```{r PEfail2}
range(SF_resample,na.rm = TRUE)
as_nc_0 <- analyze_survival(RD = data[,-1],c_range = c(20))
as_nc_0$uncertainty[[3]]
```

## (2) The cooperative case

Now, making the same comparison as in (1) for the cooperative cell line BT20 
shows the disastrous effect of ignoring the coefficient $b$, when it is 
in fact different from $1$.

```{r PEfailCoop, fig.width=5, fig.height=4}
data(CFAdata)
data <- subset.data.frame(x = CFAdata, subset = CFAdata$cell.line == "BT20")
data <- aggregate(x = data[, -1], by = list(data$`Cells seeded`), FUN = "mean", na.rm = TRUE)
PE_x <- data$`4 Gy` / data$`Cells seeded`
PE_0 <- data$`0 Gy` / data$`Cells seeded`
plot(x = rep(c(0, 1), each = length(PE_x)), 
     y = c(PE_0, PE_x), lty = 0, ylim = c(0,0.08),xlim = c(-0.1,1.1),
     xlab = "treatment", ylab = "C / S", axes = FALSE, main = "BT20")
axis(side = 1,at = c(0,1),labels = c("0 Gy","4 Gy"))
axis(side = 2,at = seq(0,0.08,0.02),las = 1)
SF_resample <- rep(PE_x, each = length(PE_0)) / rep(PE_0, times = length(PE_x))
hist(SF_resample, breaks = 100,xlim = c(0,10),xlab = "(C_4/S_4) / (C_0/S_0)",
     main = "valid PE-based SF'-values")
```

The range of the PE-based $SF'$-values does not correspond to the 
uncertainty of the $SF(20)$-values:
```{r PEfailCoop2, fig.width=6, fig.height=5}
range(SF_resample,na.rm = TRUE)
as_c_4 <- analyze_survival(RD = data[,-1],c_range = c(20))
as_c_4$uncertainty[[3]]
```

Even though the survival fraction can be accurately estimated under 
consideration of cellular cooperation 
```{r PEfailCoop2posta}
sem_SF20 <- (as_c_4$uncertainty[[3]][2]-as_c_4$uncertainty[[3]][1])/4
cat("    SF(20) : ",round(as_c_4$SF[[3]],digits = 3),"\n")
cat("SEM SF(20) : ",round(sem_SF20,digits = 3),"\n")
```
the PE-based approach fails in 
returning a trustworthy estimate of the fraction of cells losing their 
potential due to the treatment (see histogram).

In particular, the average of PE-based $SF'$ calculations does not 
asymptotically tend to an interpretable value.
In case of strong cellular cooperation, the PE-based calculated $SF'$-value is heavily 
affected by this cellular cooperation and the treatment effect of interest is
degraded to a side effect: 
```{r PEfailCoop2postb}
cat("    SF'    : ",round(mean(SF_resample,na.rm = TRUE),digits = 3),"\n")
cat("SEM SF'    : ",round(sd(SF_resample,na.rm = TRUE)/
                            sqrt(length(SF_resample)),digits = 3),"\n")
```
As a consequence, the fraction of cells keeping their clonogenic potential 
after treatment is far outside the uncertainty interval of the average 
$SF'$-value of the PE-based approach.

## Conclusion from (1) and (2)

Before calculating PE-based survival fractions, one must check whether 
there is cellular cooperation or not.

Essentially, to decide, whether you have a cooperativity issue or not, 
you need to conduct the same analysis and to generate the same data 
that is necessary to solve this issue anyways.

<!-- # Full CFAdata Set -->
<!-- ```{r Full, fig.width=13, fig.height=6} -->
<!-- data("CFAdata") -->
<!-- SF <- vector('list',7) -->
<!-- for (i in seq_along(SF)){ -->
<!--   SF[[i]] <- analyze_survival( -->
<!--     RD = subset.data.frame(x = CFAdata, -->
<!--                            subset = CFAdata$cell.line==levels(CFAdata$cell.line)[i])[,-1], -->
<!--     name = levels(CFAdata$cell.line)[i], -->
<!--     xtreat = c(0,1,2,4,6,8), -->
<!--     c_range = c(5,100)) -->
<!-- } -->
<!-- plot_sf(SF) -->
<!-- ``` -->
